Using Complex Networks for Identifying Potentially New Saline Stress Responsive Genes in Rice

### 2020-10-09 camila/report/framework.tex
references added

Next step in the workflow is to identifying communities (also called modules)
from the co-expression network structure and dynamics represented in $A$. 
The idea is to cluster genes with similar patterns of differential
expression change. Membership in these modules may overlap in 
biological contexts, because modules may be related to specific 
molecular, cellular or tissue functions and the biological components
(i.e. genes) may be involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec:HLC}).

>>>>

Next step in the workflow is to identifying communities (also called modules)
from the co-expression network structure and dynamics represented in $A$. 
The idea is to cluster genes with similar patterns of differential
expression change. Membership in these modules may overlap in 
biological contexts, because modules may be related to specific 
molecular, cellular or tissue functions and the biological components
(i.e. genes) may be involved in multiple functions. Thus, unlike WGCNA, the adjacency matrix $A$ is used to detect overlapping (rather than non-overlapping) communities, using the Hierarchical Link Clustering (HLC) algorithm (see section~\ref{sec.prelim}).

### 2020-10-09 camila/report/intro.tex
references added

Abiotic stresses are key factors that can negatively influence plant
development and productivity. They are a main cause of extensive
agricultural production losses worldwide~\cite{todo}. Soil salinity is
one of the most devastating abiotic stresses, causing reduction in the
cultivable land, crop quality, and productivity. It has been estimated
that 20\% of total cultivated and 33\% of irrigated agricultural lands
worldwide are already affected by high salinity. Moreover, due to the
human activities and natural causes, salinized areas are gradually
increasing every year and are expected to reach 50\% by the end of
year 2050~\cite{shrivastava2015soil}. Salinity tolerance and
susceptibility in plants is known to be the result of elaborated
interactions between morphological, physiological, and biochemical
processes that are regulated, in the end, by multiple genes in
different parts of a genome~\cite{reddy2017salt}. Therefore,
identifying groups of stress responsive genes may lead to crop
improvement in terms of salinity tolerance and, ultimately, contribute
solutions to the general problem of food sustainability in the years
to come.

@@

Technically, the proposed approach is both a generalization and an
extension of Weighted Gene Co-expression Network Analysis
(WGCNA)~\cite{todo2}, a widely applied workflow that has been
successfully used for identifying target genes related to diseases and
cancer in several organisms~\cite{tian2018identifying}. The general
idea behind each approach is to identify specific modules in a network
of genes after a sequence of normalization and filtering steps. The
proposed approach is considered a \textit{generalization} of WGCNA
because module detection can now recognize overlapping communities,
which may have more biological meaning given the overlapping
regulatory domains of systems that generate
co-expression~\cite{gaiteri2014beyond}. This is achieved by using
Hierarchical Link Clustering (HLC)~\cite{ahn2010link}. It is also an
\textit{extension} of WGCNA because additional steps and information
are added to the workflow: namely, some networks in the intermediate
steps are forced to be scale-free~\cite{todo3} and LASSO
regression~\cite{tibshirani1996regression} is employed to select the
most significant modules of phenotypical responses to stress.  The
advantage of using HLC as clustering method is its ability to detect
overlapping modules, since biological components are involved in
multiple functions and therefore biological communities tend to be
highly overlapping. On the other hand, LASSO is a regularized
regression technique widely used in variable selection, thanks to its
ability to obtain zero regression coefficients for the less relevant
variables~\cite{desboulets2018review}. Moreover, LASSO is especially
useful in problems where the number of variables is much larger than
the number of samples, which may be the case more often than desired.
The proposed workflow is also modular, since other module detection
and selection techniques could be used, instead HLC and LASSO,
respectively.

>>>>

Abiotic stresses are key factors that can negatively influence plant
development and productivity. They are a main cause of extensive
agricultural production losses worldwide~\cite{mesterhazy2020losses}. Soil salinity is
one of the most devastating abiotic stresses, causing reduction in the
cultivable land, crop quality, and productivity. It has been estimated
that 20\% of total cultivated and 33\% of irrigated agricultural lands
worldwide are already affected by high salinity. Moreover, due to the
human activities and natural causes, salinized areas are gradually
increasing every year and are expected to reach 50\% by the end of
year 2050~\cite{shrivastava2015soil}. Salinity tolerance and
susceptibility in plants is known to be the result of elaborated
interactions between morphological, physiological, and biochemical
processes that are regulated, in the end, by multiple genes in
different parts of a genome~\cite{reddy2017salt}. Therefore,
identifying groups of stress responsive genes may lead to crop
improvement in terms of salinity tolerance and, ultimately, contribute
solutions to the general problem of food sustainability in the years
to come.

@@

Technically, the proposed approach is both a generalization and an
extension of Weighted Gene Co-expression Network Analysis
(WGCNA)~\cite{langfelder2008wgcna}, a widely applied workflow that has been
successfully used for identifying target genes related to diseases and
cancer in several organisms~\cite{tian2018identifying}. The general
idea behind each approach is to identify specific modules in a network
of genes after a sequence of normalization and filtering steps. The
proposed approach is considered a \textit{generalization} of WGCNA
because module detection can now recognize overlapping communities,
which may have more biological meaning given the overlapping
regulatory domains of systems that generate
co-expression~\cite{gaiteri2014beyond}. This is achieved by using
Hierarchical Link Clustering (HLC)~\cite{ahn2010link}. It is also an
\textit{extension} of WGCNA because additional steps and information
are added to the workflow: namely, some networks in the intermediate
steps are forced to be scale-free~\cite{todo3} and LASSO
regression~\cite{tibshirani1996regression} is employed to select the
most significant modules of phenotypical responses to stress.  The
advantage of using HLC as clustering method is its ability to detect
overlapping modules, since biological components are involved in
multiple functions and therefore biological communities tend to be
highly overlapping. On the other hand, LASSO is a regularized
regression technique widely used in variable selection, thanks to its
ability to obtain zero regression coefficients for the less relevant
variables~\cite{desboulets2018review}. Moreover, LASSO is especially
useful in problems where the number of variables is much larger than
the number of samples, which may be the case more often than desired.
The proposed workflow is also modular, since other module detection
and selection techniques could be used, instead HLC and LASSO,
respectively.
